---
title: "Class 11 Lab Session"
author: "Patrick Tran"
format: pdf
always_allow_html: yes
---

# Section 1. Proportion og G/G in a population

Downloaded a CSV file from ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39999105-40207112;v=rs8067378;vdb=variation;vf=105535077;sample=Mexican%20Ancestry%20in%20Los%20Angeles#373531_tablePanel >

Here we read this CSV file.

```{r}
mxl <-  read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Let's now dig into this further.

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
library(plotly)
```

Let's make a boxplot

```{r}
e <- ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
e
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
ggplotly(e)
```

Sample size is 462. The median expression levels for each of the genotypes are shown in the boxplot. A/A median is 31.25. A/G median is 25.06. G/G median is 20.07.


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

The relative expression median value for A/A is greater than the median value for G/G. The SNP A/G genotype is about the average expression value of the A/A and G/G genotype.



